Estimating duration using survival analysis

### 2023-07-14 cis-perfect-testing.tex
Edits for CIS perfect testing chapter

I take a survival analysis approach to model the coarse sampling of the CIS (see \cref{perf-test:sec:problem}).
Flat priors can be unintentionally very informative, and hence weakly informative priors should be used; in addition, the prior can be used to combine the information in the CIS with the results of \cref{E-ATACCC} (see \cref{perf-test:sec:parameters-priors}).
Through a simulation study (see \cref{perf-test:sec:simulation-study}), I show that these methods perform well, and in this context are not sensitive to model or prior specification.
However, application in reality must account for misclassification bias, the subject of \cref{E-imperf-test}.

@@

The CIS infrequently tests each individual, with tests occurring either weekly or monthly (see\todo{refer to section with CIS intro}).
The infrequent testing complicates analysis of the data because the infection episode endpoints (start and end times) are known imprecisely, furthermore, infection episodes can be undetected.
This chapter develops a method to account for the infrequent testing; this section describes my approach and its background within the relevant statistical literature.

I use survival analysis for analysing this data.
Survival analysis is the area of statistics concerned with estimating the distribution of the times between events.
It is broadly applicable to many domains, such as the time-to-failure for a mechanical system or the effect of a treatment on time spent in hospital.
In the context of this thesis, the distribution to estimate is that of duration of positivity.
The \emph{initiating event} is when the duration starts, \ie becoming detectable.
The \emph{terminating event} is when the duration ends, \ie becoming no longer detectable.

This chapter uses survival analysis rather than the biomarker approach in \autoref{E-ATACCC} for computational efficiency and data limitations.
Computational efficiency is an issue because the number of random effects in the biomarker approach scales with the number of detected infection episodes.
Therefore, the biomarker approach would be computationally challenging with the large sample size in CIS.
The computational challenge is compounded by the data being stored in a trusted research environment limiting the computational power and tools available.
CIS data is limited by the infrequent sampling, meaning that most episodes have only a single positive test.
With only a single positive test, there is no information on the viral load trajectory, and hence the random effects are unlikely to be well-identified.
The biomarker approach does use more information per sample, taking into account the viral load as well as the binary test result (\ie whether the result is positive or negative), however, this is unlikely to be sufficient to overcome these problems.

The first complication is that infections episodes can be undetected, meaning no there are no positive tests associated with them.
This happens when an episode starts and ends between tests (see \cref{perf-test:fig:truncation}).
For these undetected episodes, there is no information on them meaning that we do not even know how many of them occurred.
Undetected episodes are important to account for because they are systematically shorter than detected episodes.
To see why this is the case, consider the undetected infection in the bottom of \cref{perf-test:fig:truncation}.
The infection episode depicted begins at time 33 and has a 10-day duration.
If the duration was 24 days (or longer), the individual would still have been detectable at their next test, and therefore would have been detected.
Undetected infections are a large problem in CIS because individuals are tested rarely compared to the duration of an infection episode, particularly when tested only every 28 days.
However, we do still know the testing schedule of the individual as they are in the CIS cohort.
Undetected events where we have some data on the events in which these events occur is an unusual situtation.
Far more common is that there is no data on these events, a situation known as \emph{truncation}.
A situation that is particularly relevant is \emph{left truncation}, when only events lasting a minimum amount of time are observed.
\begin{figure}
  \centering \includegraphics{cis-perfect-testing/truncation}
  \caption[Truncation in CIS data]{Episodes data in CIS is truncated, meaning that an unknown number of infections are never detected. Consider the two individuals shown here: we collect identical data for them (a series of negative tests) yet the top individual was never infected and the bottom individual was. See the main text for why this is an issue. \label{perf-test:fig:truncation}}
\end{figure}

The second complication is that the endpoints of detected infection episodes are known imprecisely.
An event, such as an endpoint, is known \emph{censored} if the time of the event can only be bounded
Censored data is present in almost all survival analyses.
An event is \emph{left censored} if it is known to occur before a certain time, \emph{right censored} if it is known to occur after a certain time, and \emph{interval censored} if it is known to occur within an interval.
%Left and right censoring can be viewed as special cases of interval censoring where the lower bound is negative infinity or the upper bound is positive infinity respectively.
Interval censoring arises in CIS because we only observe the change from being not detectable (\ie: negative) at one test, to detectable (\ie: positive) at the following test and vice versa (see \cref{perf-test:fig:double-interval-censor}).
\emph{Double censoring} means that both the originating and terminating event are censored, and is also a well-studied problem.
Doubly censored data is more challenging to handle because the initiating event time and duration must be jointly modelled~\autocite[and references therein]{liSemiparametric}.
CIS data is \emph{double interval censoring}, meaning that both the initiating and the terminating events are interval censored; approaches for double interval censoring are reviewed by \textcite{sunAnalysis,bogaertsSurvival}.
\begin{figure}
  \centering \includegraphics{cis-perfect-testing/double-interval-censor}
  \caption[Double-interval censoring in CIS data]{Episodes data in CIS is double interval censored, meaning that both the start and end of the episode are only known up to an interval. Demonstrated here by a participant who is recorded negative at time 7 to positive at time 14, bounding the start of the episode; similarly, changing from positive at time 28 to negative at time 56 bounds the end of the episode. \label{perf-test:fig:double-interval-censor}}
\end{figure}

Few studies consider the combination of both double censoring (of any type) and undetected or truncated events
This is especially true within the human biostatistical literature.
This literature though did generate general theoretical frameworks in the 1970s which include double censored and truncated case~\autocites{turnbullEmpirical}{dempsterMaximum}.
Double censored and truncated data do occur in studies estimating the time from being infected with HIV to being diagnosed with AIDS (the disease that HIV causes) such as \textcite{sunEmpirical,bacchettiNonparametric}.
These studies (and some others~\autocite{shenNonparametric}) take a conditional likelihood approach, conditioning on both the truncation (a selection effect) and the period where the infection is censored to\todo{clumsy sentence}.
They condition on the time of infection is because this information was considered to provide little information on the incubation period~\citePersonalComms{Nick Jewell}.
The period in which the infection is detected in CIS is likely to be in the period where there is frequent testing.
Therefore, conditioning on this period overstates the probability of the infection being detected.

@@

The hazard is convenient to use because it is unconstrained except in the interval [0, 1]\todo{ref where been used previously}.
Survival needs to be monotonic.

\subsection{Independent priors} \label{perf-test:sec:independent-priors}
Start with the standard assumption of independent priors on each parameter.
Each hazard, $\lambda_t$, is a (conditional) probability, and hence a Beta distribution is a natural choice.

Uninformative Beta priors, with equal parameters, on the hazard are problematic in this context.
An uninformative prior would be of the form $\lambda_t \dist \Beta(\alpha, \alpha)$.
Commonly, $\alpha$ is chosen to be $0.5$ (Jeffreys' Prior) or $1$ (a uniform prior).
Even though these priors are uninformative or flat on the hazard, they become highly informative on the survival. 
Specifically, they tend to favour shorter survival times (see \cref{perf-test:fig:flat-prior}).
An intuitive explanation of this can be drawn from the fact that these types of priors have an expected value of 0.5.
The expected survival time, $\E \left( \prod_{i=1}^{t-1} (1-\lambda_t) \right)$, is therefore equal to $0.5^{t-1}$, a quantity that declines rapidly.
This prior expresses extreme scepticism in estimates from previous studies, such as \cref{E-ATACCC} and the meta-analysis of \textcite{cevikShedding}.
\begin{figure}
  \centering \includegraphics{cis-perfect-testing/flat-prior}
  \caption[Uninformative priors for the hazard]{The density of Beta(1, 1) flat prior and Beta(0.5, 0.5) uninformative priors (left) with the prior predictive survival time (right). The implied prior on the survival time is very short when using either prior on the hazard at each time. \label{perf-test:fig:flat-prior}}
\end{figure}

@@

A second-order random walk prior encodes that the hazard should be linearly changing with some random changes at each time step.
\begin{align}
  \logit\lambda_{t+1}
  &= \logit\lambda_t + (\logit\lambda_t - \logit\lambda_{t-1}) + \sigma \epsilon_t &\text{for $t \geq 2$} \\
  &= 2\logit\lambda_t - \logit\lambda_{t-1} + \sigma \epsilon_t \\
  \epsilon_t &\dist N(0, 1) &\text{for $t \geq 2$}  \\
  \logit\lambda_2 &= \logit\lambda_1 + \epsilon_1 \\\
  \epsilon_1 &\dist N(\mu_2, \sigma_2^2) \\
  \logit \lambda_1 &\dist N(\mu_1, \sigma_1^2)
\end{align}
The hyperparameter $\sigma$ controls the standard deviation of the random walk, and there is an additional prior on $\epsilon_1$ for the initial gradient of the change in the hazard.

@@

I propose combining the information in CIS with \cref{E-ATACCC}'s analysis by using \cref{E-ATACCC}'s analysis as a prior for the CIS analysis.
This is desirable because the ATACCC study frequently samples individuals early in the infection.
Therefore, that analysis's estimate of the survival distribution in the early infection period should be reliable.
In the later infection period, the ATACCC study has fewer observations and hence the estimates are less reliable; CIS can then inform the posterior in this region.

In the combination methodology, two aspects need to be considered.
Firstly, the model structure from \cref{E-ATACCC} leads to positive correlation in the posterior estimates of the hazard, which should be propagated into this analysis.
That is, the prior used in this analysis cannot be independent across the hazards.
Secondly, the uncertainty in the estimates from \cref{E-ATACCC} should be increased for this analysis.
The additional uncertainty is because the \cref{E-ATACCC} analysis extrapolates beyond the data using strong model assumptions (see \todo{insert location}).
Furthermore, there are differences in the study design and laboratories used between the two studies (see \cref{E-intro:sec:studies}) which may mean that results do not generalise between the studies.
I form the prior for the combination in two steps.

I first approximate the \cref{E-ATACCC} posterior estimate of the hazard as a multivariate normal on the logit scale.
This can be viewed as an approximation of Markov melding~\autocite{goudieJoining}.
Using a multivariate normal, as opposed to multiple univariate distribution, ensures that the correlation between the hazards is preserved.
The logit transformation maps from the $[0, 1]$ interval to the full real line.
To estimate the parameters of the multivariate normal, I use the method of moments.
The approximation is very good (see \cref{perf-test:fig:approximate-ATACCC}).

@@

\subsection{Descriptive analyses of simulated data}

\subsection{Inference and hyperparameters}

Stan including settings

Priors on missed infections.

Survival prior hyperparameters (maybe this should just go in the previous section though?)

Criteria for assessing convergence (Rhat, ESS, divergent transitions)

@@

Points to make:

\begin{enumerate}
  \item Convergence assessment
  \item Repeated simulation showed very little variation (large sample size)
  \item Recovers true value of survival well with little difference between priors, \cref{perf-test:fig:survival-results}
  \item RW2 performs poorly considering hazard and when little information in prior very uncertain, \cref{perf-test:fig:hazard-results}; not clear why this is (\cref{perf-test:fig:hazard-pairs-results})
  \item Do I need some quantitive performance measures or is just showing plots enough?
\end{enumerate}

\begin{figure}
  \centering \includegraphics{cis-perfect-testing/survival-results}
  \caption[Comparison of survival function estimates under different priors]{Comparison of the posterior estimate of the survival distribution (pointwise median and 95\% CrI) when using different priors on the hazard. \label{perf-test:fig:survival-results}}
\end{figure}

\begin{figure}
  \centering \includegraphics{cis-perfect-testing/hazard-results}
  \caption[Comparison of hazard estimates under different priors]{Comparison of the posterior estimate of the hazard (pointwise median and 95\% CrI) when using different priors on the hazard. \label{perf-test:fig:hazard-results}}
\end{figure}

\begin{figure}
  \centering \includegraphics{cis-perfect-testing/hazard-pairs-results}
  \todo[inline]{Probably scrap this (correlation) plot}
  \caption[Correlation between hazard estimates]{Correlation between posterior estimates of the hazard when using the vague prior. \label{perf-test:fig:hazard-pairs-results}}
\end{figure}

@@

Discussion points:
\begin{enumerate}
  \item Other priors (GPs?)
  \item Inclusion of covariates: age, previous exposure status, variant.
  \item Individual vs total model
  \item Consideration of smaller sample size
\end{enumerate}

Limitations to include:
\begin{enumerate}
  \item Assumption of constant incidence
\end{enumerate}

\section{Conclusion} \label{perf-test:sec:conclusion}

>>>>

I take a Bayesian survival analysis approach to model the coarse sampling of the CIS (see \cref{perf-test:sec:problem}).
Flat priors can be unintentionally very informative, and hence weakly informative priors should be used; in addition, the prior can be used to combine the information in the CIS with the results of \cref{E-ATACCC} (see \cref{perf-test:sec:parameters-priors}).
Through a simulation study, I show that these methods perform well, and in this context are not sensitive to model or prior specification (see \cref{perf-test:sec:simulation-study}).
However, application in reality must account for false negatives, the subject of \cref{E-imperf-test}.

@@

The CIS infrequently tests each individual, with tests occurring either weekly or monthly (see\todo{refer to section with CIS intro}).
The infrequent testing complicates analysis of the data in two ways: infection episodes can be undetected and the infection episode endpoints (start and end times) are known imprecisely.
This chapter develops a method to account for the infrequent testing.
This section describes my approach and its background within the relevant statistical literature.

I analyse the data using survival analysis.
Survival analysis is the area of statistics concerned with estimating the distribution of the times between two events.
The event at the start of the time of interest is the \emph{initiating event}, and the event at the end of the time of interest is the \emph{terminating event}.
It is broadly applicable to many domains, such as the time-to-failure for a mechanical system or the effect of a treatment on time spent in hospital.
In the context of this thesis, the distribution to estimate is the duration of positivity.
Therefore, the initiating event is when the duration starts, \ie becoming detectable; and the terminating event is when the duration ends, \ie becoming no longer detectable.

\todo[inline]{I think the following paragraph is needed somewhere, to answer why I'm not doing the same thing as in ATACCC, but it feels like an aside here since the previous paragraph and the one after this one are more linked.}
Here, I use survival analysis rather than the biomarker approach taken in \autoref{E-ATACCC} for computational efficiency and data limitations.
In the biomarker approach, computational efficiency is an issue because the number of random effects scales with the number of detected infection episodes.
There are a large number of detected episodes in CIS 
Therefore, the biomarker approach would be computationally challenging with the large sample size in CIS.
The computational challenge is compounded by the data being stored in a trusted research environment which limits the computational power and tools available.
The data limitation is due to the infrequent sampling in the CIS study design, meaning that most episodes have only a single positive test.
With only a single positive test, there is no information on the viral load trajectory, and hence the random effects are unlikely to be well-identified.
While the biomarker approach does use more information per sample by taking into account the viral load as well as the binary test result (\ie whether the result is positive or negative), however, this is unlikely to be sufficient to overcome these problems.

The first complication in the analysis is that infections episodes can be undetected, meaning no there are no positive tests associated with them.
This happens when an episode starts and ends between tests (see \cref{perf-test:fig:truncation}).
For these undetected episodes, there is no information on them meaning that we do not even know how many of them occurred.
Undetected episodes are important to account for because they are systematically shorter than detected episodes.
To see why this is the case, consider the undetected infection in the bottom of \cref{perf-test:fig:truncation}.
The infection episode depicted begins at time 33 and has a 10-day duration.
If the duration was 24 days (or longer), the individual would still have been detectable at their next test, and therefore would have been detected.
Undetected infections are a large problem in CIS because individuals are tested rarely compared to the duration of an infection episode, particularly when tested only every 28 days.
However, we do still know the testing schedule of the individual as they are in the CIS cohort.
Undetected events where we have some data on the events in which these events occur is an unusual situtation.
Far more common is that there is no data on these events, a situation known as \emph{truncation}.
A situation that is particularly relevant is \emph{left truncation}, when only events lasting a minimum amount of time are observed.
\begin{figure}
  \centering \includegraphics{cis-perfect-testing/truncation}
  \caption[Truncation in CIS data]{Episodes data in CIS is truncated, meaning that an unknown number of infections are never detected. Consider the two individuals shown here: we collect identical data for them (a series of negative tests) yet the top individual was never infected and the bottom individual was. See the main text for why this is an issue. \label{perf-test:fig:truncation}}
\end{figure}

The second complication is that the endpoints of detected infection episodes are known imprecisely.
An event, such as an endpoint, is known as \emph{censored} if the time of the event can only be bounded.
Censored data is present in almost all survival analyses.
An event is \emph{left censored} if it is known to occur before a certain time, \emph{right censored} if it is known to occur after a certain time, and \emph{interval censored} if it is known to occur within an interval.
%Left and right censoring can be viewed as special cases of interval censoring where the lower bound is negative infinity or the upper bound is positive infinity respectively.
Interval censoring arises in CIS because we only observe the change from being not detectable (\ie: negative) at one test, to detectable (\ie: positive) at the following test and vice versa (see \cref{perf-test:fig:double-interval-censor}).
If both the initiaing and terminating events are censored, then the event is \emph{doubly censored}.
Doubly censored data is more challenging to handle because the initiating event time and duration must be jointly modelled~\autocite[and references therein]{liSemiparametric}.
In CIS data both the initiating and the terminating events are interval censored.
Therefore, the CIS data is \emph{double interval censored}; approaches for double interval censoring are reviewed by \textcite{sunAnalysis,bogaertsSurvival}.
\begin{figure}
  \centering \includegraphics{cis-perfect-testing/double-interval-censor}
  \caption[Double-interval censoring in CIS data]{Episodes data in CIS is double interval censored, meaning that both the start and end of the episode are only known up to an interval. Demonstrated here by a participant who is recorded negative at time 7 to positive at time 14, bounding the start of the episode; similarly, changing from positive at time 28 to negative at time 56 bounds the end of the episode. \label{perf-test:fig:double-interval-censor}}
\end{figure}

Few studies consider the combination of both double interval censoring and undetected or truncated events
This is especially true within the human biostatistical literature.
While this literature does contain theoretical frameworks which include double censored and truncated case~\autocites{turnbullEmpirical}{dempsterMaximum}, they were normally applied when the terminating event was either uncensored or right censored~\autocite{sunEmpirical,bacchettiNonparametric}.
The approach of these studies (and elsewhere~\autocite{shenNonparametric}) was to take a conditional likelihood approach, conditioning on both the truncation (a selection effect) and the period in which the initiating event is known to occur.
They justify conditioning on the time of the initiating event by arguing that this information provides little information on the distribution being estimated~\citePersonalComms{Nick Jewell}.
The period in which the infection is detected in CIS is likely to be in the period where there is frequent testing.
Therefore, conditioning on this period overstates the probability of the infection being detected.

@@

The hazard is convenient to use because it is unconstrained except in the interval [0, 1].
The [0, 1] interval is easily be mapped to the unconstrained reals using a logit transformation (where $\logit(x) = \log(x/(1-x))$).
Conversely, the survival function, $S_\theta(t)$, needs to be monotonically decreasing.
Therefore, there is a constraint on $S_\theta(t)$ which depends on $S_\theta(t-1)$.
This complicates the inference process and cannot be removed through a simple transformation.

The rest of this section considers appropriate priors for the hazard in this context.

\subsection{Independent priors} \label{perf-test:sec:independent-priors}
Take the standard assumption of independent priors on each parameter.
Each hazard, $\lambda_t$, is a (conditional) probability, and hence a Beta distribution is a natural choice.

Uninformative Beta priors on the hazard are problematic in this context.
An uninformative prior would be of the form $\lambda_t \dist \Beta(\alpha, \alpha)$.
Commonly, $\alpha$ is chosen to be $0.5$ (Jeffreys' Prior) or $1$ (a uniform prior).
Even though these priors are uninformative on the hazard, they become highly informative on the survival. 
Specifically, they tend to favour shorter survival times (see \cref{perf-test:fig:flat-prior}).
An intuitive explanation of this can be drawn from the fact that these types of priors have an expected value of 0.5.
The expected survival time, $\E \left( \prod_{i=1}^{t-1} (1-\lambda_t) \right)$, is therefore equal to $0.5^{t-1}$, a quantity that declines rapidly.
This prior expresses extreme scepticism in estimates from previous studies, such as \cref{E-ATACCC} and the meta-analysis of \textcite{cevikShedding}.
\begin{figure}
  \centering \includegraphics{cis-perfect-testing/flat-prior}
  \caption[Uninformative priors for the hazard]{The density of Beta(1, 1) flat prior and Beta(0.5, 0.5) uninformative priors (left) with the prior predictive survival time (right). The implied prior on the survival time is very short when using either prior on the hazard at each time. \label{perf-test:fig:flat-prior}}
\end{figure}

@@

A second-order random walk prior encodes that the hazard should be linearly changing with some random changes at each time step.
Specifically, the difference between $\lambda_t$ to $\lambda_{t+1}$ is the difference between $\lambda_{t-1}$ and $\lambda_t$ plus some Gaussian-distributed noise.
Formally:
\begin{align}
  \logit\lambda_{t+1}
  &= \logit\lambda_t + (\logit\lambda_t - \logit\lambda_{t-1}) + \sigma \epsilon_t &\text{for $t \geq 2$} \\
  &= 2\logit\lambda_t - \logit\lambda_{t-1} + \sigma \epsilon_t \\
  \epsilon_t &\dist N(0, 1) &\text{for $t \geq 2$}  \\
  \logit\lambda_2 &= \logit\lambda_1 + \epsilon_1 \\\
  \epsilon_1 &\dist N(\mu_{\epsilon_1}, \sigma_{\epsilon_1}^2) \\
  \logit \lambda_1 &\dist N(\mu_{\lambda_1}, \sigma_{\lambda_1}^2) \\
  \sigma &\dist \Exponential(1/\mu_\sigma).
\end{align}
The hyperparmeters $\mu_{\lambda_1}$ and $\sigma_{\lambda_1}$ specify the prior on the hazard at time 1; $\mu_{\epsilon_1}$ and $\sigma_{\epsilon_1}$ the prior on the initial gradient; and $\mu_\sigma$ the smoothness of the random walk.
I use $\mu_{\lambda_1} = -17.5$, $\sigma_{\lambda_1} = 6$, $\mu_\epsilon = 1.09$, $\sigma_\epsilon = 0.03$, and $\mu_\epsilon = 0.1$ based on the analysis in \cref{E-ATACCC}.

@@

I propose combining the information from the CIS data with \cref{E-ATACCC}'s analysis by using \cref{E-ATACCC}'s analysis as a prior for the CIS analysis.
This is desirable because the ATACCC study frequently samples individuals early in the infection.
Therefore, that analysis's estimate of the survival distribution in the early infection period should be reliable.
In the later infection period, the ATACCC study has fewer observations and hence the estimates are less reliable; CIS can then inform the posterior in this region.

In the combination methodology, two aspects need to be considered.
Firstly, the model structure from \cref{E-ATACCC} leads to positive correlation in the posterior estimates of the hazard, which should be propagated into this analysis.
That is, the prior used in this analysis cannot be independent across the hazards.
Secondly, the uncertainty in the estimates from \cref{E-ATACCC} should be increased for this analysis.
The additional uncertainty is because the \cref{E-ATACCC} analysis extrapolates beyond the data using strong model assumptions (see \todo{reference discussion on extrapolation in ATACCC chapter}).
Furthermore, there are differences in the study design and laboratories used between the two studies (see \cref{E-intro:sec:studies}) which may mean that results do not generalise between the studies.
I form the prior for the combination in two steps.

I first approximate the \cref{E-ATACCC} posterior estimate of the hazard as a multivariate normal on the logit scale.
This can be viewed as an approximation of Markov melding~\autocite{goudieJoining}.
Using a multivariate normal, as opposed to multiple univariate distribution, ensures that the correlation between the hazards is preserved.
The logit transformation maps from the $[0, 1]$ interval to the full real line.
To estimate the parameters of the multivariate normal, I use the method of moments.
The approximation is very good (see \cref{perf-test:fig:approximate-ATACCC}).
\todo[inline]{Add figure showing the approximation is good}

@@

\subsection{Implementation}

Simulation was performed in R~4.2.0~\autocite{R4-2-0} using tidyverse~2.0.0~\autocite{tidyverse} and targets~1.1.3~\autocite{targetsPackage}.

Inference was implemented in Stan via RStan~2.21.8~\autocite{rstan2-21-8} using default settings.
Convergence was assessed using Rhat and ESS as described in \cref{E-ATACCC:sec:convergence}, and all runs checked for divergent transitions.

The prior on the total number of infections was TBD\todo{fill in once decided on total vs individual model}.

A framework for performing CIS-like simulations is available at \url{https://github.com/joshuablake/cisSimulation}.
A framework for fitting the class of models described here and in \cref{E-imperf-test} is available at \url{https://github.com/joshuablake/cisDurationModel}.
Code utilising these frameworks to perform the analysis described in this chapter is available at \url{https://github.com/joshuablake/cisRuns}.

@@

All analyses converged satisfactorily\todo{put numbers once confirmed exact analyses to use}.
Repeating simulations showed very little variation in the results, due to the large sample size in CIS; therefore, only results from a single simulation are shown.

The true value of the survival function is very well recovered with little difference between priors (see \cref{perf-test:fig:survival-results}).

The independent prior has the fewest assumptions and least prior information.
Therefore, it leads to posterior estimates that are most uncertain.

The second-order random walk is the only prior that smooths the hazard throughout the entire infection.
This leads to a near-constant hazard, and hence exponential-like survival, in areas where there is little information (long infections).

\begin{figure}
  \centering \includegraphics{cis-perfect-testing/survival-results}
  \caption[Comparison of survival function estimates under different priors]{Comparison of the posterior estimate of the survival distribution (pointwise median and 95\% CrI) when using different priors on the hazard. Results using each of the three different priors discussed in \cref{perf-test:sec:parameters-priors} are shown. \label{perf-test:fig:survival-results}}
\end{figure}

\begin{figure}
  \centering \includegraphics{cis-perfect-testing/hazard-results}
  \caption[Comparison of hazard estimates under different priors]{Comparison of the posterior estimate of the hazard (pointwise median and 95\% CrI) when using different priors on the hazard. Results using each of the three different priors discussed in \cref{perf-test:sec:parameters-priors} are shown. \label{perf-test:fig:hazard-results}}
\end{figure}

@@

This chapter shows that the model I propose can recover the true survival function well, is not sensitive to the prior used for the survival distribution, nor requires information on the number of infections that were undetected.
However, the problem becomes substantially harder if false negatives are present, as we shall see in the next chapter.
Unfortunately, false negatives do occur in the CIS data.

Various extensions could be worth exploring.
Most importantly, covariates (\eg age, immunity from vaccination and/or previous infection, or variant) could be included.
Immunity in particular affects the duration substantially, although age and variant may also make a contribution~\autocites{hakkiOnset}{russellWithinhost}.
Covariates can be included with a variety of model, the simplest being a proportional hazards model.

The form of priors used for the survival function could also be extended.
In particular, the form of smoothing chosen here is very simple and could be replaced with a more flexible model, such as a Gaussian Process for additional flexibility and smoothing and different time-scales~\autocite{saulGaussian}.

The prior would likely be more important if the number of detected infections was smaller.
Therefore, exploring its impact would be important for designing future surveys.
A particular application would be estimation in the early stages of an epidemic of a novel pathogen.
Here, little would be known about the pathogen, but gathering information with robust estimates of uncertainty is important to inform the response.

I have assumed constant and independent incidence (a priori).
In this case, a sensitivity analysis (not shown) showed minimal impact on survival estimates.
Furthermore, I chose a period where the incidence was approximately constant.
However, this assumption could be important in other contexts, especially if the incidence is changing rapidly.
The assumption of constant incidence can lead to biased estimates\todo{cite}.

\section{Conclusion} \label{perf-test:sec:conclusion}
This chapter has shown the feasibility of estimating the survival function from a study with a CIS-like testing schedule.
While the infrequent testing schedule is not ideal, appropriate modelling can recover the true survival function well.
Next, we will see that this model is inadequate when false negatives are present, such as they are within the CIS data.
